MS Technical Paper: \\ Placement Algorithms for Heterogeneous FPGAs

### 2025-05-08 latex/paper/sections/07-rapidwright_api.tex
carry chain traversal

\item \textbf{Key Packages:} 
    RapidWright revolves around three primary data model packages:
    \begin{enumerate}
    \item \emph{edif} -- Represents the logical netlist in an abstracted EDIF-like structure.
    \item \emph{design} -- Contains data structures for the physical implementation (Cells, Nets, Sites, BELs, etc.).
    \item \emph{device} -- Provides a database of the target FPGA architecture (e.g., Site coordinates, Tile definitions, routing resources).
    \end{enumerate}

>>>>

\item \textbf{Key Packages:} 
    RapidWright revolves around three primary data model packages:
    \begin{enumerate}
    \item \texttt{edif} -- Represents the logical netlist in an abstracted EDIF-like structure.
    \item \texttt{design} -- Contains data structures for the physical implementation (Cells, Nets, Sites, BELs, etc.).
    \item \texttt{device} -- Provides a database of the target FPGA architecture (e.g., Site coordinates, Tile definitions, routing resources).
    \end{enumerate}

### 2025-05-08 latex/paper/sections/08-netlist.tex
carry chain traversal

@@

\newpage

\end{multicols}
{
    \raggedright
    \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/top_level.png}
    \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_0.png}
    \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_1.png}
    \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_2.png}
    \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_3.png}
    \captionof{figure}{A simple HDL design with module hierarchy.}
    \label{fig:hierarchical_design}
}
\vspace{0.5cm}
{
    \centering
    \includegraphics[valign=c, width=11.5cm]{figures/netlist_synth/hier_netlist.png}
    \includegraphics[valign=c, width=6cm]{figures/netlist_synth/hier_graph.png}
    \captionof{figure}{
        \textbf{Left:} A hierarchical netlist consisting of LUTs and FFs.
        \textbf{Right:} The cell hierarchy graph.
    }
    \label{fig:hier_netlist}
}
\vspace{0.5cm}
{
    \centering
    \includegraphics[valign=c, width=13cm]{figures/netlist_synth/flat_netlist.png}
    \includegraphics[valign=c, width=4cm]{figures/netlist_synth/flat_graph.png}
    \captionof{figure}{
        \textbf{Left:} A flattened netlist consisting of LUTs and FFs.
        \textbf{Right:} The flattened cell hierarchy graph.
    }
    \label{fig:flat_netlist}
}
\newpage
\begin{multicols}{2}

@@

\begin{lstlisting}[language=java, caption={Netlist extraction and traversal}, label={lst:netlist_extract}]
Design design = Design.readCheckpoint("synth.dcp")
EDIFNetlist = design.getNetlist();

@@

\newcolumn

>>>>

\newpage
\end{multicols}
{
    \raggedright
    \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/top_level.png}
    \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_0.png}
    \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_1.png}
    \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_2.png}
    \includegraphics[valign=t, scale=0.3]{figures/netlist_synth/module_3.png}
    \captionof{figure}{A simple HDL design with module hierarchy.}
    \label{fig:hierarchical_design}
}
\vspace{0.5cm}
{
    \centering
    \includegraphics[valign=c, width=11.5cm]{figures/netlist_synth/hier_netlist.png}
    \includegraphics[valign=c, width=6cm]{figures/netlist_synth/hier_graph.png}
    \captionof{figure}{
        \textbf{Left:} A hierarchical netlist consisting of LUTs and FFs.
        \textbf{Right:} The cell hierarchy tree.
    }
    \label{fig:hier_netlist}
}
\vspace{0.5cm}
{
    \centering
    \includegraphics[valign=c, width=11.5cm]{figures/netlist_synth/flat_netlist.png}
    \includegraphics[valign=c, width=4cm]{figures/netlist_synth/flat_graph.png}
    \captionof{figure}{
        \textbf{Left:} A flattened netlist consisting of LUTs and FFs.
        \textbf{Right:} The flattened cell hierarchy tree.
    }
    \label{fig:flat_netlist}
}
\newpage
\begin{multicols}{2}

@@

@@

\begin{lstlisting}[language=java, caption={Netlist extraction and traversal}, label={lst:netlist_extract}]
Design design = Design.readCheckpoint("synth.dcp")
EDIFNetlist netlist = design.getNetlist();

@@

\begin{lstlisting}[caption={Code Printout}]
Net: dout[0]
	Port: I0, Cell: dout_i_1, CellType: LUT4
	Port: Q, Cell: m0/dout_reg, CellType: FDRE
Net: q_1
	Port: I2, Cell: dout_i_1, CellType: LUT4
	Port: Q, Cell: m0/q_1_reg, CellType: FDRE
	Port: I5, Cell: q_1_i_1, CellType: LUT6
Net: ce
	Port: I1, Cell: dout_i_1, CellType: LUT4
	Port: I1, Cell: dout_i_1__0, CellType: LUT6
	Port: I1, Cell: dout_i_1__1, CellType: LUT6
	Port: I3, Cell: q_1_i_1, CellType: LUT6
Net: dout[1]
	Port: I0, Cell: dout_i_1__0, CellType: LUT6
	Port: Q, Cell: m1/m2/dout_reg, CellType: FDRE
Net: dout_i_1__1_n_0
	Port: O, Cell: dout_i_1__1, CellType: LUT6
	Port: D, Cell: m1/m3/dout_reg, CellType: FDRE
Net: clk
	Port: C, Cell: m0/dout_reg, CellType: FDRE
	Port: C, Cell: m0/q_1_reg, CellType: FDRE
	Port: C, Cell: m1/m2/dout_reg, CellType: FDRE
	Port: C, Cell: m1/m3/dout_reg, CellType: FDRE
Net: dout[2]
	Port: I0, Cell: dout_i_1__1, CellType: LUT6
	Port: Q, Cell: m1/m3/dout_reg, CellType: FDRE
Net: <const0>
	Port: G, Cell: GND, CellType: GND
	Port: R, Cell: m0/dout_reg, CellType: FDRE
	Port: R, Cell: m0/q_1_reg, CellType: FDRE
	Port: R, Cell: m1/m2/dout_reg, CellType: FDRE
	Port: R, Cell: m1/m3/dout_reg, CellType: FDRE
Net: <const1>
	Port: P, Cell: VCC, CellType: VCC
	Port: CE, Cell: m0/dout_reg, CellType: FDRE
	Port: CE, Cell: m0/q_1_reg, CellType: FDRE
	Port: CE, Cell: m1/m2/dout_reg, CellType: FDRE
	Port: CE, Cell: m1/m3/dout_reg, CellType: FDRE
Net: dout_i_1__0_n_0
	Port: O, Cell: dout_i_1__0, CellType: LUT6
	Port: D, Cell: m1/m2/dout_reg, CellType: FDRE
Net: rst
	Port: I3, Cell: dout_i_1, CellType: LUT4
	Port: I5, Cell: dout_i_1__0, CellType: LUT6
	Port: I5, Cell: dout_i_1__1, CellType: LUT6
	Port: I4, Cell: q_1_i_1, CellType: LUT6
Net: dinc
	Port: I2, Cell: dout_i_1__0, CellType: LUT6
	Port: I2, Cell: dout_i_1__1, CellType: LUT6
	Port: I2, Cell: q_1_i_1, CellType: LUT6
Net: dinb
	Port: I3, Cell: dout_i_1__0, CellType: LUT6
	Port: I4, Cell: dout_i_1__1, CellType: LUT6
	Port: I0, Cell: q_1_i_1, CellType: LUT6
Net: dina
	Port: I4, Cell: dout_i_1__0, CellType: LUT6
	Port: I3, Cell: dout_i_1__1, CellType: LUT6
	Port: I1, Cell: q_1_i_1, CellType: LUT6
Net: q_1_i_1_n_0
	Port: D, Cell: m0/q_1_reg, CellType: FDRE
	Port: O, Cell: q_1_i_1, CellType: LUT6
Net: dout_i_1_n_0
	Port: O, Cell: dout_i_1, CellType: LUT4
	Port: D, Cell: m0/dout_reg, CellType: FDRE
\end{lstlisting}

### 2025-05-08 latex/paper/sections/10-simulated_annealing.tex
carry chain traversal

\section{Simulated Annealing}
\label{sec:simulated_annealing}
With a basic understanding of FPGA architecture, design placement, and RapidWright, we have all the necessary pieces to implement our SA placer. 
Here we outline in detail each substage of our implementation: PrePacking, Packing, and Placement. 
Shown in Figure \ref{fig:edif_design_device} shows an overview of the placement workflow.

@@

    The raw post-synthesis netlist only tells us the list of nets and the cell ports that they connect to. 
    It does not report the presence of any macro cell structures (CARRY4 chains, etc.). 
    Thus, we must traverse the netlist to detect these cell structures and store that structure information in a class we will call \texttt{PrepackedDesign}.

@@

    Utilizing all 8 LUT-FF lanes in a SLICE can help reduce device area utilization and minimize wirelength, but \emph{too much} logic density in an area can contribute to general routing congestion. 
    Furthermore, attempting to fill all 8 lanes in a SLICE requires some careful consideration of LUT BEL constraints. 
    The LUT BELs in a SLICE, as shown in Figure \ref{fig:slicel}, the LUTs appear to be interleaved.

    Show a block diagram of a LUT6 comprising of two LUT5s.

    Recall that a LUT can accommodate one 6-input boolean function or two 5-input boolean functions sharing the same inputs or any two 3-input or less boolean functions regardless of shared inputs. 
    A LUT6 Cell (a LUT with 6-inputs) will actually occupy two lanes in a SLICEL, rendering one of the FF BELs in either lane ineligible for another LUT-FF pair. 
    The LUT BELs

>>>>

\section{Placement}
\label{sec:simulated_annealing}
With a basic understanding of FPGA architecture, design placement, and RapidWright, we have all the necessary pieces to implement our SA placer. 
Here we outline in detail each substage of our implementation: PrePacking, Packing, and Placement. 
Shown in Figure \ref{fig:edif_design_device} is an overview of the placement workflow. 
Figure \ref{fig:substages} shows the data structures of RapidWright objects that are populated at each stage.

@@

The raw EDIF netlist only tells us the list of nets and the cell ports that they connect to. 
It does not report the presence of any macro cell structures (CARRY4 chains, etc.). 
Thus, we must traverse the netlist to detect these cell structures and store that structure information in a class we will call \texttt{PrepackedDesign}.

@@

Utilizing all 8 LUT-FF lanes in a SLICE can help reduce device area utilization and minimize wirelength, but \emph{too much} logic density in an area can contribute to general routing congestion. 
Furthermore, attempting to fill all 8 lanes in a SLICE requires meticulous adherence to conditional LUT constraints. 
Recall that a LUT can accommodate one 6-input boolean function or two 5-input boolean functions sharing the same inputs or any two 3-input or less boolean functions regardless of shared inputs. 
A LUT6 Cell (a LUT with 6-inputs) will actually occupy two LUT-FF lanes in a SLICEL, rendering one of the FF BELs in either lane ineligible for another LUT-FF pair. 
To simplify the problem space, we will only fill up to four LUT-FF lanes in any given SLICE.

\begin{lstlisting}[language=java, caption={Netlist extraction and traversal}, label={lst:netlist_extract}]
Design design = Design.readCheckpoint("synth.dcp")
EDIFNetlist netlist = design.getNetlist();
List<EDIFCellInst> ecis = netlist.getAllLeafCellInstances();

// Select only the carry cells.
for (EDIFCellInst eci : ecis) {
    if (eci.getCellName() == "CARRY4")
        carryCells.add(eci);
}

// Find and remove carry chains until the list is empty
while (!carryCells.isEmpty()) {
    // Arbitrarily set "currentCell" pointer to a cell in the list
    EDIFCellInst currentCell = carryCells.get(0);

    // Find this carry chain anchor.
    // Traverse the Carry-In (CI) to Carry-Out (CO) nets.
    // Anchor is found when net on the CI Port is Ground.
    while (true) {
        // Access the CI port on this cell.
        EDIFPortInst sinkPort = currentCell.getPortInst("CI");
        // Access the net on this CI port.
        EDIFNet net = sinkPort.getNet();
        if (net.isGND())
            // Found this chain anchor!
            break;
        // Get all ports on this net.
        List<EDIFPortInst> netPorts = net.getPortInsts();
        for (EDIFPortInst netPort : netPorts) {
            // Access the port belonging to another carry cell.
            EDIFCellInst sourceCell = netPort.getCellInst();
            if (sourceCell.getCellName() == "CARRY4") {
                // Move the "currentCell" pointer
                currentCell = sourceCell;
            }
        }
    }

    // Now we have the chain anchor as currentCell.
    // Now traverse in the opposite direction to find the chain tail.
    // Tail is found when the CO Port is null.
    // Collect the chain cells into an ordred list.
    List<EDIFCellInst> currentChain = new ArrayList<>();
    while (true) {
        EDIFPortInst sourcePort = currentCell.getPortInst("CO[3]");
        if (sourcePort == null)
            // Found this chain's tail!
            break;
        EDIFNet net = sourcePort.getNet();
        List<EDIFPortInst> sinkPorts = net.getPortInsts();
        for (EDIFPortInst sinkPort : sinkPorts) {
            EDIFCellInst sinkCell = sinkPort.getCellInst();
            if (sinkCell.getCellName() == "CARRY4") {
                currentCell = sinkCell;
                // Add the cell to the chain list.
                currentChain.add(currentCell);
            }
        }
    }

    // Add currentChain to the list of chains
    carryChains.add(currentChain);
}
\end{lstlisting}
